### 2018-06-14 express2018/part2.tex
Other changes mainly in part2

The final relationship between traces and weak transitions is stated
and proved in the following lemma: (here $us$ as single variable means
a list of actions)
\begin{lemma}
A weak transition $E\overset{u}{\Rightarrow}E'$ is a just trace with non
empty action list: 1) without any visible label, if $u = \tau$, or 2)
$u$ is the unique label in the list, if $u \neq \tau$.
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{E} \HOLTokenWeakTransBegin\HOLFreeVar{u}\HOLTokenWeakTransEnd \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenEquiv{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{us}.
       \HOLConst{TRACE} \HOLFreeVar{E} \HOLBoundVar{us} \HOLFreeVar{E\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenNeg{}}\HOLConst{NULL} \HOLBoundVar{us} \HOLSymConst{\HOLTokenConj{}} \HOLKeyword{if} \HOLFreeVar{u} \HOLSymConst{=} \HOLSymConst{\ensuremath{\tau}} \HOLKeyword{then} \HOLConst{NO_LABEL} \HOLBoundVar{us}
       \HOLKeyword{else} \HOLConst{UNIQUE_LABEL} \HOLFreeVar{u} \HOLBoundVar{us}\hfill{[WEAK_TRANS_AND_TRACE]}
\end{alltt}
\end{lemma}

To finish the proof of Lemma~\ref{l:ruptocon}, we use the  above
result to freely switch between weak transitions and traces
during the induction argument.
% , which either keep its
% length or become shorter after passing weak bisimilations, then we
% used above lemma again to convert the traces back to weak transitions.
Below is the formally proved version of
Theorem~\ref{t:contraBisimulationU}: (Here \texttt{WGS} stands for weakly guarded context without direct
sums)
\begin{alltt}
UNIQUE_SOLUTION_OF_CONTRACTIONS:
\HOLTokenTurnstile{} \HOLConst{WGS} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{Q}
\end{alltt}

@@

The formal proof of ``unique solution of rooted contractions theorem''
(Theorem~\ref{t:rcontraBisimulationU}) has the
same initial proof steps as Theorem~\ref{t:contraBisimulationU} plus a
little more steps to handle the rooted bisimilarity at conclusion. (In the thesis work, the conclusion of this theorem
is weak bisimilarity, and the proof is exactly the same as Unique
solution of contractions theorem. Now we got a stronger conclusion
using observational congruence, and the previous version becomes a
trivial collorary of the current version.) The
two proofs are quite similar, mostly because the only property we need
from (rooted) contraction is its precongruence. Once we have proved
the precongruence of rooted contracion, we can naturally use the
normal version of weakly guarded expressions with direct sums
included. Below is the formally proved version of Theorem~\ref{t:rcontraBisimulationU}:
\begin{alltt}
UNIQUE_SOLUTION_OF_ROOTED_CONTRACTIONS:
\HOLTokenTurnstile{} \HOLConst{WG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenObsContracts} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenObsContracts} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{Q}
\end{alltt}

>>>>

The final relationship between traces and weak transitions is stated
and proved in the following lemma: (here $acts$ as a single variable means
a list of actions)
\begin{lemma}
A weak transition $P\overset{u}{\Rightarrow}P'$ is a just trace with non
empty action list: 1) without any visible label, if $u = \tau$, or 2)
$u$ is the unique label in the list, if $u \neq \tau$.
\begin{alltt}
\HOLTokenTurnstile{} \HOLFreeVar{P} \HOLTokenWeakTransBegin\HOLFreeVar{u}\HOLTokenWeakTransEnd \HOLFreeVar{P\sp{\prime}} \HOLSymConst{\HOLTokenEquiv{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{acts}.
       \HOLConst{TRACE} \HOLFreeVar{P} \HOLBoundVar{acts} \HOLFreeVar{P\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenNeg{}}\HOLConst{NULL} \HOLBoundVar{acts} \HOLSymConst{\HOLTokenConj{}}
       \HOLKeyword{if} \HOLFreeVar{u} \HOLSymConst{=} \HOLSymConst{\ensuremath{\tau}} \HOLKeyword{then} \HOLConst{NO_LABEL} \HOLBoundVar{acts} \HOLKeyword{else} \HOLConst{UNIQUE_LABEL} \HOLFreeVar{u} \HOLBoundVar{acts}\hfill{[WEAK_TRANS_AND_TRACE]}
\end{alltt}
\end{lemma}

To finish the proof of Lemma~\ref{l:ruptocon}, we have used the  above
result to freely switch between weak transitions and traces
during the induction argument.
% , which either keep its
% length or become shorter after passing weak bisimilations, then we
% used above lemma again to convert the traces back to weak transitions.
Below is the formally proved version of
Theorem~\ref{t:contraBisimulationU}: (Here \texttt{WGS} stands for weakly guarded context without direct
sums)
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WGS} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{Q}
\hfill{[UNIQUE_SOLUTION_OF_CONTRACTIONS]}
\end{alltt}

@@

The formal proof of ``unique solution of rooted contractions theorem''
(Theorem~\ref{t:rcontraBisimulationU}) has the
same initial proof steps as Theorem~\ref{t:contraBisimulationU} plus a
little more steps to handle the rooted bisimilarity at conclusion. The
two proofs are quite similar, mostly because the only property we need
from (rooted) contraction is its precongruence. Once we have proved
the precongruence of rooted contracion, we can naturally use the
normal version of weakly guarded expressions with direct sums
included. Below is the formally verified version of
Theorem~\ref{t:rcontraBisimulationU}:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenObsContracts} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenObsContracts} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{Q}
\hfill{[UNIQUE_SOLUTION_OF_ROOTED_CONTRACTIONS]}
\end{alltt}

See how similar it is with Milner's original ``unique solution of
equations'' theorem for strong bisimilarity ($\sim$) in which the same
weak guardness (\texttt{WG}) is required:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{WG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenStrongEQ} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenStrongEQ} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenStrongEQ} \HOLBoundVar{Q}\hfill{[STRONG_UNIQUE_SOLUTION]}
\end{alltt}

Or our Theorem~\ref{t:rcontraBisimulationU} can be seen as a more
applicable version of Milner's ``unique solution of
equations'' theorem for rooted bisimilarity ($\rapprox$), which has more
restrictions on equations:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{SG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{SEQ} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenObsCongr} \HOLBoundVar{Q}
\hfill{[OBS_UNIQUE_SOLUTION]}
\end{alltt}

### 2018-06-14 express2018/related.tex
Other changes mainly in part2

Monica Nesi formalized both pure CCS \cite{Nesi:1992ve} and Value-Passing CCS
\cite{Nesi:2017wo} in old versions of HOL theorem prover, but the
related proof scripts were not publically available.\footnote{Part of
 the formalization of pure CCS can be found at \url{https://github.com/binghe/informatica-public/tree/master/CCS2/CCS-Nesi}.}
She was mainly focusing on writing decision procedures (ML functions) which
automatically prove theorems showning bisimilarity of CCS
processes.%
% \footnote{The work is ported (by the 2nd author) from
% HOL88 to latest HOL4, with small improvements on existing work, then
% added the theory of context and (pre)congruence, coarsest
% (pre)congruence contained in bisimilarity, bisimulation up to,
% expansion, contraction, and unique solution of
% equations/expansions/contractions et al, these additions have doubled
% the size of original work.}

Beside Nesi, there're three other CCS formalisations. One
is done in Isabelle/HOL (based on Nominal datatypes) in 2010, by Jesper Bengtson as part of his PhD
thesis project \cite{bengtson2010formalising}, available in Isabelle's Archive of Formal
Proofs (AFP) \footnote{\url{https://www.isa-afp.org/entries/CCS.html}}. He has formalized
classic properties for strong/weak equivalence and observational
congruence, however the CCS grammar doesn't include process constants
(or recursion operators), and the proved theorems also look quite
different with their forms in textbooks. On the other side, Jesper
Bengtson and Joachim Parrow made great progress on pi-calculus formalization and
proved that the algebraic axiomatization of bisimulation
equivalence in the pi-calculus is sound and complete. \cite{bengtson2007completeness}
The other formalisation is based on Coq, done by Solange
Coupet-Grimal in 1995, which is part of the
``coq-contribs'' archive (Coq's official user contributed code)
\footnote{\url{https://github.com/coq-contribs/ccs}}. This is actually
a formalisation of transition systems without any algebraic operator,
and covers the very basic properties of strong/weak equivalence and observational
congruence. The last one was done by Chaudhuri et
al. \cite{chaudhuri2014formalization} using Abella, this work
particularily focus on bisimulation up to techniques while the CCS
type doesn't have restriction, recursion and relabelling operators.

>>>>

Monica Nesi formalized both pure CCS \cite{Nesi:1992ve} and Value-Passing CCS
\cite{Nesi:2017wo} in old versions of HOL theorem prover, but the
related proof scripts were not publically available.\footnote{Part of
 the formalization of pure CCS can be found at \url{https://github.com/binghe/informatica-public/tree/master/CCS2/CCS-Nesi}.}
She was mainly focusing on writing decision procedures (ML functions) which
automatically prove theorems showing the bisimilarity of two given CCS
processes.%
% \footnote{The work is ported (by the 2nd author) from
% HOL88 to latest HOL4, with small improvements on existing work, then
% added the theory of context and (pre)congruence, coarsest
% (pre)congruence contained in bisimilarity, bisimulation up to,
% expansion, contraction, and unique solution of
% equations/expansions/contractions et al, these additions have doubled
% the size of original work.}

Beside Nesi, there exist other CCS formalisations. Jesper Bengtson
formalized CCS, $\pi$-calculi and $\psi$-calculi in his PhD
thesis \cite{bengtson2010formalising} using Isabelle/HOL and its
nominal logic. For CCS, he has formalized
basic properties for strong/weak equivalence and observational
congruence, however the CCS syntax doesn't have constants
(or recursion operator), and the proved theorems also look quite
different with their forms in textbooks.
On the other side, Jesper
Bengtson and Joachim Parrow made great progress on $\pi$-calculus formalization and
proved that the algebraic axiomatization of bisimulation
equivalence in the $\pi$-calculi is sound and complete. \cite{bengtson2007completeness}
The other formalisation is based on Coq, done by Solange
Coupet-Grimal in 1995, which is part of the
``coq-contribs'' archive (Coq's official user contributed code)
\footnote{\url{https://github.com/coq-contribs/ccs}}. This is actually
a formalisation of transition systems without any algebraic operator,
and covers the very basic properties of strong/weak equivalence and observational
congruence. The last one was done by Chaudhuri et
al. \cite{chaudhuri2014formalization} using Abella, this work
particularily focus on bisimulation up to techniques while the CCS
type doesn't have restriction, recursion and relabelling operators.
